The 2- and 3-point surface correlation functions calculations: from novel exact continuous approach to improving methodology for discrete images

### 2023-05-17 paper.tex
A few minor improvements

Another function studied in this paper is surface-surface-surface function
$F_{sss}$. It is defined as follows:
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = \langle M_S(\bm{x}) M_S(\bm{x} + \bm{r_1})
  M_S(\bm{x} + \bm{r_2}) \rangle
  \label{eq:fsss}
\end{equation}
This function is defined for media with dimensionality 3 or higher and is a
physical quantity inversely proportional to volume. SI units for $F_{sss}$ are
$m^{-3}$, but relative units can be used in practice. For three-dimensional media
$F_{sss}$ is a function of six scalar arguments:
$\bm{r_1} = (r_{1x}, r_{1y}, r_{1z})$ and $\bm{r_2} = (r_{2x}, r_{2y}, r_{2z})$.
However, if the medium under consideration is isotropic, only three scalar
arguments $r_1 = |\bm{r_1}|$, $r_2 = |\bm{r_2}|$ and $r_3 = |\bm{r_1} - \bm{r_2}|$
are required. The last argument can be replaced with an angle $\theta$ between
vectors $\bm{r_1}$ and $\bm{r_2}$:
\begin{equation}
  F_{sss}(r_1, r_2, r_3) = F_{sss}(r_1, r_2, \theta) = F_{sss}(\bm{r_1}, \bm{r_2})
\end{equation}

@@

\section{Computation of surface functions for digital images}
\label{sec:algo}
Here we briefly describe the algorithm for computation of surface-surface,
surface-void and surface-surface-surface correlation functions for digital
images of porous media \cite{Samarin}. This discrete algorithm can calculate
surface correlation functions for all possible directions and correlation
lengths. It consists of two steps: edge detection and computation of
autocorrelation. Next, we present some improvements in this approach that later
will be verified with the help of exact continuous method.

@@

For surface-surface-surface function $F_{sss}$ the previously described approach
can also be used, but it results in inefficient memory usage because for
three-dimensional case $F_{sss}$ is a function of six scalar coordinates, which
leads to a very large output array. Therefore, instead we use a somewhat slower,
but more memory efficient, approach that computes $F_{sss}$ at a specified point
calculating a convolution using its definition:
\begin{algorithmic}[1]
  \Procedure{surfsurfsurf}{$A$, $\bm{r_1}$, $\bm{r_2}$}
  \State $A' \gets |H*A|$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $A'_1 \gets shift(A', \bm{r_1})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{1i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $A'_2 \gets shift(A', \bm{r_2})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{2i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $M \gets A' \cdot A'_1 \cdot A'_2$
  \Comment{Multiplicate three arrays element-wise}
  \State \textbf{return} $\sum_{i} M_{i} / N$
  \Comment{Sum all elements in $M$ and divide by a number of elements in $A$}
  \EndProcedure
\end{algorithmic}
This algorithm can be vectorized and therefore benefits from the usage of
GPUs. Shift function in this algorithm must be a circular shift which
``rotates'' content of the array, if periodic boundary conditions are
desired. For our purpose we use $F_{sss}$ function in the form
$F_{sss}(r_1, r_2, \theta)$ and fix $\theta = \pi/2$. This leads us to a
sampling pattern shown on \cref{fig:Fsss-pattern}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/pattern.png}
  \caption[]{Sampling pattern used for calculation of $F_{sss}$ in our
    implementation is a right triangle with fixed catets $r_1$ and $r_2$.
    Patterns can lie in planes $XY$ (as on the picture), $XZ$ or $YZ$. This
    pattern is moved over the input which results in all possible distinct
    configurations. Black dots depict elements of $A'$.}
  \label{fig:Fsss-pattern}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/experiment-setup.png}
  \caption[]{Left, center: images $Q_1$ and $Q_2$ used in the demonstration of
    approximation for $1 / \sin \phi$. Right: Interfaces between phases are
    line segments which intersect each other at angle $\phi$. Shaded area is
    where $|(H*Q_1\cdot(H*Q_2)|$ has non-zero values. Sum of all values in this
    area approximates $1 / \sin \phi$.}
  \label{fig:experiment-setup}
\end{figure}
Consider two 2D images like the ones on \cref{fig:experiment-setup}. Boundaries
between black and white areas cross each other at the angle $\phi$. We apply an
edge detection filter to each image, multiply results element-wise and sum all
elements of the product. We want the resulting value to approximate
$1/\sin \phi$. As can be seen on \cref{fig:filter-comparison} $H'$ does this job
better than $H$. Why this is important will be explained in the next section.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/filter-comparison.png}
  \caption[]{Estimation of $1/\sin\phi$ by filters $H$ and $H'$ for binary
    images on \cref{fig:experiment-setup}.}
  \label{fig:filter-comparison}
\end{figure}

>>>>

Another function studied in this paper is three-point surface-surface-surface
correlation function $F_{sss}$. It is defined as follows:
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = \langle M_S(\bm{x}) M_S(\bm{x} + \bm{r_1})
  M_S(\bm{x} + \bm{r_2}) \rangle
  \label{eq:fsss}
\end{equation}
This function is defined for media with dimensionality 3 or higher and is a
physical quantity inversely proportional to volume. SI units for $F_{sss}$ are
$m^{-3}$, but relative units can be used in practice. For three-dimensional media
$F_{sss}$ is a function of six scalar arguments:
$\bm{r_1} = (r_{1x}, r_{1y}, r_{1z})$ and $\bm{r_2} = (r_{2x}, r_{2y}, r_{2z})$.
However, if the medium under consideration is isotropic, only three scalar
arguments $r_1 = |\bm{r_1}|$, $r_2 = |\bm{r_2}|$ and $r_3 = |\bm{r_1} - \bm{r_2}|$
are required. The last argument can be replaced with an angle $\theta$ between
vectors $\bm{r_1}$ and $\bm{r_2}$:
\begin{equation}
  F_{sss}(\bm{r_1}, \bm{r_2}) = F_{sss}(r_1, r_2, r_3) = F_{sss}(r_1, r_2, \theta)
\end{equation}

@@

\section{Computation of surface functions for digital images}
\label{sec:algo}
Here we briefly describe the algorithm for computation of $F_{ss}$, $F_{sv}$ and
$F_{sss}$ correlation functions for digital images of porous media
\cite{Samarin}. This discrete algorithm calculates surface correlation functions
for all possible directions and correlation lengths. It consists of two steps:
edge detection and computation of autocorrelation. Next, we present some
improvements in this approach that later will be verified with the help of exact
continuous method.

@@

For surface-surface-surface function $F_{sss}$ the previously described approach
can also be used, but it results in inefficient memory usage because for
three-dimensional case $F_{sss}$ is a function of six scalar coordinates, which
leads to a very large output array. Therefore, instead we use a somewhat slower,
but more memory efficient, approach that computes $F_{sss}$ at a specified point
calculating a convolution using its definition:
\begin{algorithmic}[1]
  \Procedure{surfsurfsurf}{$A$, $\bm{r_1}$, $\bm{r_2}$}
  \State $A' \gets |H*A|$
  \Comment{$A'$ becomes an interface of $A$.}
  \State $A'_1 \gets shift(A', \bm{r_1})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{1i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $A'_2 \gets shift(A', \bm{r_2})$
  \Comment{Shift $A$ in $i$-th dimension by $r_{2i}$ for all $i \in \overline{1, ndim(A)}$}
  \State $M \gets A' \cdot A'_1 \cdot A'_2$
  \Comment{Multiplicate three arrays element-wise}
  \State \textbf{return} $\sum_{i} M_{i} / N$
  \Comment{Sum all elements in $M$ and divide by a number of elements in $A$}
  \EndProcedure
\end{algorithmic}
This algorithm can be vectorized and therefore benefits from the usage of
GPUs. Shift function in this algorithm must be a circular shift which
``rotates'' content of the array, if periodic boundary conditions are
desired. For our purpose we represent $F_{sss}$ function in the form
$F_{sss}(r_1, r_2, \theta)$ and fix $\theta = \pi/2$. This leads us to a
sampling pattern shown on \cref{fig:Fsss-pattern}
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/pattern.png}
  \caption[]{Sampling pattern used for calculation of $F_{sss}$ in our
    implementation is a right triangle with fixed catets $r_1$ and $r_2$.
    Patterns can lie in planes $XY$ (as on the picture), $XZ$ or $YZ$. This
    pattern is moved over the input which results in all possible distinct
    configurations. Black dots depict elements of $A'$.}
  \label{fig:Fsss-pattern}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/experiment-setup.png}
  \caption[]{Left, center: images $Q_1$ and $Q_2$ used in the demonstration of
    approximation for $1 / \sin \phi$. Right: Interfaces between phases are
    line segments which intersect each other at angle $\phi$. Shaded area is
    where $|(H*Q_1\cdot(H*Q_2)|$ has non-zero values. Sum of all values in this
    area approximates $1 / \sin \phi$.}
  \label{fig:experiment-setup}
\end{figure}
Consider two 2D images like the ones on \cref{fig:experiment-setup}. Boundaries
between black and white areas cross each other at the angle $\phi$. We apply an
edge detection filter to each image, multiply results element-wise and sum all
elements of the product. We want the resulting value to approximate
$1/\sin \phi$. As can be seen on \cref{fig:filter-comparison} $H'$ does this job
better than $H$. Why this is important will be explained in \cref{sec:fss-2d}.
\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{images/filter-comparison.png}
  \caption[]{Estimation of $1/\sin\phi$ by filters $H$ and $H'$ for binary
    images on \cref{fig:experiment-setup}.}
  \label{fig:filter-comparison}
\end{figure}
